% Acknowlegements

First and foremost, I would like to thank my advisor, Dr. Roman Makarevich for his guidance throughout this project, which would not have been possible without his continuous support and feedback.  I would also like to acknowledge the efforts of the other members of my committee, Dr. William Bristow, Dr. Mark Conde, and Dr. Hui Zhang for offering new ideas and perspectives.  Additionally, I truely appreciate the support and encouragement I received from the entire physics department at UAF, particularly Dr. David Neuman, Dr. Renate Wackerbauer, Dr. Curt Szuberla, Dr. Chung-Sang Ng, Dr. Peter Delamere, Ellen Craig and Jeanie Talbot, as well as my fellow grad students, Victoriya, Christina, Jason, Robin, Joe, Nathan, Soma, Bishwa, Vitaliy, and John for both the homework and coding assistance and companionship over the last several years.  I would like to extend a particularly special thank you to Flora Grabowska and all the staff of the Keith B. Mather Libray for providing invaluable assistance in retrieving a great number of old, obscure, and difficult to locate sources for my research.  

To my family, thank you all for supporting me for years, particularly my Mom, for helping me become a much better writer than I would have been otherwise, my Dad, for always picking up the phone for my programming emergencies, and my brother, Henry, for giving me a friendly voice of reason when I really needed it.  Additionally, I would like to thank my friends back East, Katie, Ari, Bonnie, Heather, Jayne, and Stephanie for supporting me over these last few years despite being 4000 miles away, and to Laura for always brightening my spirits and making my life feel a little less hectic.  Finally, a huge thank you to Allison, Rosie, Emma, and Kinley on the homefront for their tireless effort in keeping me sane, healthy, and happy throughout this whole process.  I couldn't have done it without you guys.  Thank you.